Simulating Generalized Linear Models

### 2024-01-31 paper/paper.Rmd
updating

---
title             : "Ordinal regression models made easy. A tutorial on parameter interpretation, data simulation, and power analysis."
shorttitle        : "Ordinal regression models made easy"
author: 
  - name          : "Filippo Gambarota"
    affiliation   : "1"
    corresponding : yes    # Define only one corresponding author
    address       : "Via Venezia 8, 35131, Padova (Italy)"
    email         : "filippo.gambarota@unipd.it"
    role:         # Contributorship roles (e.g., CRediT, https://casrai.org/credit/)
      - "Conceptualization"
      - "Writing - Original Draft Preparation"
      - "Writing - Review & Editing"
      - "Methodology"
      - "Software"
  - name          : "Gianmarco Altoè"
    affiliation   : "1"
    role:
      - "Conceptualization"
      - "Writing - Review & Editing"
      - "Supervision"
affiliation:
  - id            : "1"
    institution   : "Department of Developmental Psychology and Socialization, University of Padova, Italy"
abstract: |
    Ordinal data are widespread in psychology such as Likert items, ratings, or generic ordered variables. These variables are usually analyzed using metric models (e.g., standard linear regression) with important drawbacks in terms of statistical inference (reduced power and increased type-1 error) and prediction. One possible reason for not using ordinal regression models could be difficulty in understanding parameters or conducting a power analysis. The tutorial aims to present ordinal regression models using a simulation-based approach. Firstly, we introduced the general model highlighting crucial components and assumptions. Then we explained how to interpret parameters for a logit and probit model. Then we proposed two ways for simulating data as a function of predictors showing a 2x2 interaction with categorical predictors and the interaction between a numeric and categorical predictor. Finally, we showed an example of power analysis using simulations that can be easily extended to complex models with multiple predictors. The tutorial is supported by a collection of custom R functions developed to simulate and understand ordinal regression models. The code to reproduce the proposed simulation, the custom R functions, and additional examples of ordinal regression models can be found on the online Open Science Framework repository ([https://osf.io/93h5j](https://osf.io/93h5j/?view_only=33f262ee28374a88b935f0a25cefdb34)).

keywords          : "ordinal regression, monte carlo simulations, power analysis"
wordcount         : "6987"
bibliography:
  - "packages.bib"
  - "`r filor::fil()$bib`"
floatsintext      : yes
linenumbers       : no
draft             : no
mask              : no
figurelist        : no
tablelist         : no
footnotelist      : no
params:
  floatplaceh: true
  preprint: true
classoption       : "man"
output: 
  papaja::apa6_pdf:
    latex_engine: xelatex
    header_includes:
      - \usepackage{booktabs, caption, longtable, colortbl, array, hhline, mathbf}
  papaja::apa6_docx: default
---

@@

```{r tab-dataset-example-docx, eval = which_output("docx")}
papaja::apa_table(ss, caption = filor::get_caption())
```

@@

In this tutorial we presented the CM model in terms of model fitting, parameters interpretation and data simulation. The basic model can be expanded in several ways. When assuming PO or considering only a location effect is limiting, including scale (*location-scale*), thresholds (*location-shift*) effects or relaxing the PO assumption can help in modeling complex phenomenon. The supplementary materials ([https://osf.io/93h5j](https://osf.io/93h5j/?view_only=33f262ee28374a88b935f0a25cefdb34)) contains some examples of these complex models. Furtermore, with clustered data (e.g., multiple items/repetitions for the same participant), a random-effect CM model can be fitted including random intercepts and slopes.

>>>>

---
title             : "Ordinal regression models made easy. A tutorial on parameter interpretation, data simulation, and power analysis."
shorttitle        : "Ordinal regression models made easy"
author: 
  - name          : "Filippo Gambarota"
    affiliation   : "1"
    corresponding : yes    # Define only one corresponding author
    address       : "Via Venezia 8, 35131, Padova (Italy)"
    email         : "filippo.gambarota@unipd.it"
    role:         # Contributorship roles (e.g., CRediT, https://casrai.org/credit/)
      - "Conceptualization"
      - "Writing - Original Draft Preparation"
      - "Writing - Review & Editing"
      - "Methodology"
      - "Software"
  - name          : "Gianmarco Altoè"
    affiliation   : "1"
    role:
      - "Conceptualization"
      - "Writing - Review & Editing"
      - "Supervision"
affiliation:
  - id            : "1"
    institution   : "Department of Developmental Psychology and Socialization, University of Padova, Italy"
abstract: |
    Ordinal data are widespread in psychology such as Likert items, ratings, or generic ordered variables. These variables are usually analyzed using metric models (e.g., standard linear regression) with important drawbacks in terms of statistical inference (reduced power and increased type-1 error) and prediction. One possible reason for not using ordinal regression models could be difficulty in understanding parameters or conducting a power analysis. The tutorial aims to present ordinal regression models using a simulation-based approach. Firstly, we introduced the general model highlighting crucial components and assumptions. Then we explained how to interpret parameters for a logit and probit model. Then we proposed two ways for simulating data as a function of predictors showing a 2x2 interaction with categorical predictors and the interaction between a numeric and categorical predictor. Finally, we showed an example of power analysis using simulations that can be easily extended to complex models with multiple predictors. The tutorial is supported by a collection of custom R functions developed to simulate and understand ordinal regression models. The code to reproduce the proposed simulation, the custom R functions, and additional examples of ordinal regression models can be found on the online Open Science Framework repository ([https://osf.io/93h5j](https://osf.io/93h5j/?view_only=33f262ee28374a88b935f0a25cefdb34)).
keywords          : "ordinal regression, monte carlo simulations, power analysis"
wordcount         : "6987"
bibliography:
  - "packages.bib"
  - "`r filor::fil()$bib`"
floatsintext      : yes
linenumbers       : no
draft             : no
mask              : no
figurelist        : no
tablelist         : no
footnotelist      : no
params:
  floatplaceh: true
  preprint: true
  submission: true
classoption       : "man"
output: 
  papaja::apa6_docx: default
  papaja::apa6_pdf:
    latex_engine: xelatex
    header_includes:
      - \usepackage{booktabs, caption, longtable, colortbl, array, hhline, mathbf}
---

@@

```{r tab-dataset-example-docx, eval = which_output("docx")}
papaja::apa_table(ss, caption = filor::get_caption(), escape = FALSE)
```

@@

In this tutorial we presented the CM model in terms of model fitting, parameters interpretation and data simulation. The basic model can be expanded in several ways. When assuming PO or considering only a location effect is limiting, including scale (*location-scale*), thresholds (*location-shift*) effects or relaxing the PO assumption can help in modeling complex phenomenon. The supplementary materials ([https://osf.io/93h5j](https://osf.io/93h5j/?view_only=33f262ee28374a88b935f0a25cefdb34)) contains some examples of these complex models. Furthermore, with clustered data (e.g., multiple items/repetitions for the same participant), a random-effect CM model can be fitted including random intercepts and slopes.

```{r, child = 'ethical-section.Rmd'}
```

### 2024-01-31 paper/paper.tex
updating

In this tutorial we presented the CM model in terms of model fitting, parameters interpretation and data simulation. The basic model can be expanded in several ways. When assuming PO or considering only a location effect is limiting, including scale (\emph{location-scale}), thresholds (\emph{location-shift}) effects or relaxing the PO assumption can help in modeling complex phenomenon. The supplementary materials (\href{https://osf.io/93h5j/?view_only=33f262ee28374a88b935f0a25cefdb34}{https://osf.io/93h5j}) contains some examples of these complex models. Furtermore, with clustered data (e.g., multiple items/repetitions for the same participant), a random-effect CM model can be fitted including random intercepts and slopes.

>>>>

In this tutorial we presented the CM model in terms of model fitting, parameters interpretation and data simulation. The basic model can be expanded in several ways. When assuming PO or considering only a location effect is limiting, including scale (\emph{location-scale}), thresholds (\emph{location-shift}) effects or relaxing the PO assumption can help in modeling complex phenomenon. The supplementary materials (\href{https://osf.io/93h5j/?view_only=33f262ee28374a88b935f0a25cefdb34}{https://osf.io/93h5j}) contains some examples of these complex models. Furthermore, with clustered data (e.g., multiple items/repetitions for the same participant), a random-effect CM model can be fitted including random intercepts and slopes.

\section{Ethical Compliance Section}\label{ethical-compliance-section}

\subsubsection{Funding}\label{funding}

The authors have no funding to disclose.

\subsubsection{Compliance with Ethical Standards}\label{compliance-with-ethical-standards}

Compliance with Ethical Standards: Not applicable

\subsubsection{Conflicts of Interest}\label{conflicts-of-interest}

The authors declare they have no conflict of interest

### 2024-01-31 supplementary/location-scale-models.qmd
updating

Where $X\zeta$ is the linear predictor for the scale of the distribution. By default for both the *logit* and *probit* model the scale is fixed to 1. On scale-location models we put predictors on both parameters. Given that the scale cannot be negative we use a log link function $\eta = \text{log}(X\zeta)$. As suggested by @Tutz2017-de location-scale models can be considered as a more parsimonious approach compared to partially or completely relaxing the proportional odds assumption. Allowing the scale to be different as a function of the predictor create more modelling flexibility. Furthermore, two groups could be theoretically different only in the scale of the latent distribution with a similar location. In this example, the only way to capture group differences is by including a scale effect. The @fig-scale-effect-example depict the impact of having different scales between two groups on the ordinal probabilities.

```{r}
#| label: fig-scale-effect-example
cat_latent_plot(location = c(0, 1), scale = c(1, 2), prob0 = rep(1/4, 4), link = "logit", plot = FALSE) |> 
  cowplot::plot_grid(plotlist = _, nrow = 1)
```

@@

The location-scale model can be simulated using the `sim_ord_latent()` function and providing the predictors for the scale parameters. Given the `log` link function, predictors are provided on the `log` scale. For example, we simulate the effect of a binary variable $x$ representing two independent groups predicting the $k = 5$ response. We simulate a *location* effect of $\beta_1 = 0.5$ (in *probit* scale) and $\zeta_1 = \text{log}(2) = 0.70$. The first group has a $\sigma = 1$ and the second group has $\sigma = 2$. Again we simulate that the baseline probabilities are uniform for the first group.

```{r}
k <- 5  # number of options (1)
n <- 1e5 # number of observations (1)
b1 <- 0.5 # beta1, the shift in the latent distribution (2)
z1 <- log(2) # zeta1, the change in the scale
probs0 <- rep(1/k, k) # probabilities when x = 0 (3)
alphas <- prob_to_alpha(probs0, link = "probit") # get true thresholds from probabilities (3)
dat <- data.frame(x = rep(c(0, 1), each = n/2))
dat <- sim_ord_latent(~x, scale = ~x, beta = b1, zeta = z1, prob0 = probs0, data = dat, link = "probit")
fit <- clm(y ~ x, scale = ~ x, data = dat, link = "probit")
```

The table (put ref) reports the simulation results.

```{r}
#| eval: false
truth <- c(alphas, b1, z1)
names(truth) <- c(names(fit$alpha), names(fit$beta), names(fit$zeta))
#clm_table(fit, truth)
```

>>>>

Where $X\zeta$ is the linear predictor for the scale of the distribution. By default for both the *logit* and *probit* model the scale is fixed to 1. On scale-location models we put predictors on both parameters. Given that the scale cannot be negative we use a log link function $\eta = \text{log}(X\zeta)$. As suggested by @Tutz2017-de location-scale models can be considered as a more parsimonious approach compared to partially or completely relaxing the proportional odds assumption. Allowing the scale to be different as a function of the predictor create more modelling flexibility. Furthermore, two groups could be theoretically different only in the scale of the latent distribution with a similar location. In this example, the only way to capture group differences is by including a scale effect. The @fig-scale-effect-example depict the impact of having different scales between two groups on the ordinal probabilities but the same location. Clearly, there is no location effect and the two groups are predicted to be the same considering only the location effect.

```{r}
#| label: fig-scale-effect-example
cat_latent_plot(location = c(0, 0), scale = c(1, 2), prob0 = rep(1/4, 4), link = "logit", plot = FALSE) |> 
  cowplot::plot_grid(plotlist = _, nrow = 1)
```

@@

The location-scale model can be simulated using the `sim_ord_latent()` function and providing the predictors for the scale parameters. Given the `log` link function, predictors are provided on the `log` scale. For example, we simulate the effect of a binary variable $x$ representing two independent groups predicting the $k = 5$ response. We simulate a *location* effect of $\beta_1 = 0.5$ (in *probit* scale) and $\zeta_1 = \text{log}(2) = 0.70$. The first group has a $\sigma = 1$ and the second group has $\sigma = 2$. Again we simulate that the baseline probabilities are uniform for the first group. The $\log$ link function (and $e$ as inverse link function) is used to make sure that the variance is always positive. If the first group has a scale of 1 and the second group a scale of 2, the $\zeta_1 = \log(\frac{s_2}{s_1}) = \log(\frac{2}{1}) \approx 0.7$. Thus the two groups have a log difference in terms of scale of 0.7 (or a ratio of 2).

```{r}
#| echo: true
k <- 5  # number of options
n <- 1e5 # number of observations
b1 <- 0.5 # beta1, the shift in the latent distribution
z1 <- log(2) # zeta1, the change in the scale
probs0 <- rep(1/k, k) # probabilities when x = 0
alphas <- prob_to_alpha(probs0, link = "probit") # get true thresholds from probabilities
dat <- data.frame(x = rep(c(0, 1), each = n/2))
dat <- sim_ord_latent(~x, scale = ~x, beta = b1, zeta = z1, prob0 = probs0, data = dat, link = "probit")
fit <- clm(y ~ x, scale = ~ x, data = dat, link = "probit")
summary(fit)
```

### 2024-01-31 supplementary/ordinal-notes.qmd
updating

```{r}
#| fig-cap: Simulated data using the latent variable approach. The dotted lines are the thresholds $\\alpha$
n <- 1e3
x <- runif(1e3)
B <- 3
ys_p <- x * B + rnorm(n)
ys_n <- x * -B + rnorm(n)
y_p <- findInterval(ys_p, alpha) + 1
y_n <- findInterval(ys_n, alpha) + 1

>>>>

```{r}
#| fig-cap: Simulated data using the latent variable approach. The dotted lines are the thresholds $\alpha$
n <- 1e3
x <- runif(1e3)
B <- 3
ys_p <- x * B + rnorm(n)
ys_n <- x * -B + rnorm(n)
y_p <- findInterval(ys_p, alpha) + 1
y_n <- findInterval(ys_n, alpha) + 1

### 2024-01-31 supplementary/simulating-glm.qmd
major update
updating

library(tidyverse)
```

@@

The curve depict the true probability of success for each value of $x$. To include the random error we need to sample the observed values from a Binomial distribution using the vector of probabilities `p`. We sample a vector of 0 and 1.

```{r}
ns <- length(x) # number of subjects
y <- rbinom(ns, 1, p)
head(y)

>>>>

library(tidyverse)
library(cowplot)
library(ordinal)
devtools::load_all()

knitr::opts_chunk$set(echo = TRUE,
                      dev = "svg",
                      warning = FALSE,
                      message = FALSE)
mtheme <- function(){
  theme_minimal(15)
}
theme_set(mtheme())
```

@@

The curve depicts the true probability of success for each value of $x$. To include the random error we need to sample the observed values from a Binomial distribution using the vector of probabilities `p`. We sample a vector of 0 and 1.

```{r}
ns <- length(x) # number of subjects
y <- rbinom(ns, 1, p)
summary(p)
